Theoretical aspects in gravity modeling

### 2018-04-20 manuscript/letter.tex
start sec anomalous quantities in gravity modeling

Consider a TCS (Figure \ref{fig:TCS}) with origin at a point $P$ with
coordinates $(X_{P}, Y_{P}, Z_{P})$ referred to the GCS (Figure \ref{fig:fig1}).
In this case, the relationship between the coordinates $(X, Y, Z)$ of a 
point in the GCS and the coordinates $(x, y, z)$ of the same point in the 
TCS is given by:
\begin{equation}
\begin{bmatrix}
x \\
y \\
z 
\end{bmatrix} =
\mitbf{R}^{\top} \begin{bmatrix}
X - X_{P} \\
Y - Y_{P} \\
Z - Z_{P}
\end{bmatrix} \: ,
\label{eq:GCS2TCS}
\end{equation}
where $\mitbf{R}$ is an orthogonal matrix whose columns
are defined by the unit vectors $\hat{\mitbf{u}}$, $\hat{\mitbf{v}}$ and
$\hat{\mitbf{w}}$ (equation \ref{eq:unit-vectors}) as follows:
\begin{equation}
\mitbf{R} = 
\begin{bmatrix}
\hat{\mitbf{v}} & \hat{\mitbf{w}} & -\hat{\mitbf{u}}
\end{bmatrix} \: .
\label{eq:R}
\end{equation}

\section{Basic concepts}

It is worth noting that, by definition, 
the centrifugal component of the normal gravity field is
equal to the centrifugal component of the Earth's gravity
field if they are evaluated at the same point.
Then, the differences between the gravity vector
(corrected from non-gravitational effects) 
and the normal gravity vector, at the same point, represents a purely 
gravitational and consequently harmonic disturbing field.
This gravitational disturbance 
is caused by contrasts between the actual internal 
density distribution of the Earth and the internal density 
distribution of the normal Earth.
In applied geophysics, these density differences are generally 
called \textit{anomalous masses} (e.g., \citeauthor{hammer1945}, 
\citeyear{hammer1945}; \citeauthor{lafehr1965}, \citeyear{lafehr1965}),
\textit{density anomalies} (e.g., \citeauthor{forsberg1984}, \citeyear{forsberg1984})
or \textit{gravity sources} (e.g., \citeauthor{blakely1996}, 
\citeyear{blakely1996}). Here, we opted for using the last term.

The difference between the observed gravity and the
normal gravity, at the same point, is called \textit{gravity disturbance}
\citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005}.
Notice that the gravity disturbance is not equivalent to the
magnitude of the difference between the gravity vector
and the normal gravity vector at the same point 
\citep{barthelmes2013, sanso_sideris2013}.
As properly pointed out by \citet{hackney-featherstone2003},
the gravity disturbance is a very-well established quantity in geodesy,
but appears to be less well known in geophysics.

@@

\begin{figure}
%    \vspace{5.5cm}
    \includegraphics{figures/cartesian-geodetic-systems.png}
    \caption{Schematic representation of the Geocentric Cartesian System (GCS) and the Geocentric Geodetic System (GGS). The GCS has the $Z$-axis coincident with the mean Earth's rotational axis,
    the $X$-axis pointing to the Greenwich meridian and the $Y$-axis
    directed so as to obtain a right-handed system. The GGS is      defined by an oblate ellipsoid with semi-minor axis $b$, 
    coincident with the $Z$-axis of GCS, and a semi-major
    axis $a$. In this coordinate system, the position of a point is
    determined by the geometric height $h$, the geodetic latitude 
    $\varphi$ and longitude $\lambda$. 
    The Earth's center of mass is represented 
    by $O$, $P$ represents a point $(h_{P}, \varphi_{P}, \lambda_{P})$ and $P^{\prime}$ its projection onto the plane $XY$ 
    (Equatorial plane). The plane containing $O$, $P$ and 
    $P^{\prime}$ is represented in gray in (a) and (b). 
    The unit vectors $\hat{\mitbf{u}}_{P}$, $\hat{\mitbf{v}}_{P}$ and 
    $\hat{\mitbf{w}}_{P}$ define three mutually orthogonal 
    directions at $P$ (equation \ref{eq:unit-vectors}).
    In (b), $Q$ represents a point $(h_{Q}, \varphi_{Q}, \lambda_{Q})$,
    which is the projection of $P$ onto the reference ellipsoid, at
    the same latitude and longitude ($\varphi_{Q} = \varphi_{P}$ and $\lambda_{Q} = \lambda_{P}$).}
  \label{fig:GCS-GGS}
\end{figure}

\begin{figure}
    \includegraphics{figures/surfaces.png}
    \caption{text text text text}
  \label{fig:surfaces}
\end{figure}

\begin{figure}
    \includegraphics{figures/local-system.png}
    \caption{(a) Schematic representation of the gravity vector
    $\mitbf{g}_{P}$, normal gravity vector $\mitbf{\gamma}_{P}$,
    disturbing gravitational attraction $\Delta \mitbf{g}_{P}$
    (equation \ref{eq:gravity-vector}), unit vector 
    $\hat{\mitbf{u}}_{P}$ (equation \ref{eq:unit-vectors}) and
    the normal gravity vector $\mitbf{\gamma}_{Q}$ at the point $Q$ on the reference ellipsoid.
    At $P$, the normal gravity vector $\mitbf{\gamma}_{P}$ 
    is opposite to the unit vector $\hat{\mitbf{u}}_{P}$ and, at $Q$, the normal gravity $\mitbf{\gamma}_{Q}$ is opposite
    to the unit vector $\hat{\mitbf{u}}_{Q}$. These unit vectors are
    close to each other.
    (b) Schematic representation of a topocentric Cartesian 
    coordinate system with origin at a point $P$. The axes $x$
    and $y$ are parallel to the unit vectors $\hat{\mitbf{v}}_{P}$
    and $\hat{\mitbf{w}}_{P}$ (equation \ref{eq:unit-vectors} and
    Figure \ref{fig:GCS-GGS}), respectively. The $z$ axis is opposite
    to the unit vector $\hat{\mitbf{u}}_{P}$ (equation 
    \ref{eq:unit-vectors} and Figure \ref{fig:GCS-GGS}) and points downward.}
  \label{fig:TCS}
\end{figure}

>>>>

Consider a TCS (Figure \ref{fig:TCS}) with origin at a point $P$ with
coordinates $(X_{P}, Y_{P}, Z_{P})$ referred to the GCS (Figure \ref{fig:GCS-GGS}).
In this case, the relationship between the coordinates $(X, Y, Z)$ of a 
point in the GCS and the coordinates $(x, y, z)$ of the same point in the 
TCS is given by:
\begin{equation}
\begin{bmatrix}
x \\
y \\
z 
\end{bmatrix} =
\mitbf{R}^{\top} \begin{bmatrix}
X - X_{P} \\
Y - Y_{P} \\
Z - Z_{P}
\end{bmatrix} \: ,
\label{eq:GCS2TCS}
\end{equation}
where $\mitbf{R}$ is an orthogonal matrix whose columns
are defined by the unit vectors $\hat{\mitbf{u}}$, $\hat{\mitbf{v}}$ and
$\hat{\mitbf{w}}$ (equation \ref{eq:unit-vectors}) as follows:
\begin{equation}
\mitbf{R} = 
\begin{bmatrix}
\hat{\mitbf{v}} & \hat{\mitbf{w}} & -\hat{\mitbf{u}}
\end{bmatrix} \: .
\label{eq:R}
\end{equation}

\section{Anomalous quantities in gravity modeling}

Let $\mitbf{g}_{P}$ and $\mitbf{\gamma}_{P}$ be, respectively, the 
gravity vector (corrected from non-gravitational effects due to vehicle 
motion and time variations such as Earth tides, instrumental drift and 
barometric pressure changes, for example) and the normal gravity vector 
at a point $P$. The difference between these two vectors, at the same 
point $P$, defines a quantity called \textit{gravity disturbance 
vector}, which is given by:
\begin{equation}
\delta \mitbf{g}_{P} = 
\mitbf{g}_{P} - \mitbf{\gamma}_{P} \: .
\label{eq:gravity-disturbance-vector}
\end{equation}
By definition, the centrifugal component of the normal gravity vector 
is equal to the centrifugal component of the gravity vector if they 
are evaluated at the same point.
Then, the gravity disturbance vector $\delta \mitbf{g}_{P}$ represents 
a purely gravitational (and consequently harmonic) quantity, which is 
caused by contrasts between the actual internal 
density distribution of the Earth and the (unknown) internal density 
distribution of the normal Earth.
In applied geophysics, these density differences are generally 
called \textit{anomalous masses} (e.g., \citeauthor{hammer1945}, 
\citeyear{hammer1945}; \citeauthor{lafehr1965}, \citeyear{lafehr1965}),
\textit{density anomalies} (e.g., \citeauthor{forsberg1984}, 
\citeyear{forsberg1984}) or \textit{gravity sources} (e.g., 
\citeauthor{blakely1996}, \citeyear{blakely1996}). Here, we opted for 
using the last term.
Figure \ref{fig:surfaces} illustrates the gravity vector 
$\mitbf{g}_{P}$, normal gravity vector $\mitbf{\gamma}_{P}$ and 
the gravity disturbance vector $\delta \mitbf{g}_{P}$ at a point $P$ 
located on the Earth's surface.

The difference between the gravity $g_{P}$ and the normal gravity 
$\gamma_{P}$, at the same point $P$, is called \textit{gravity 
disturbance} \citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005} 
and can be represented as follows:
\begin{equation}
\delta g_{P} = g_{P} - \gamma_{P} \: .
\label{eq:gravity-disturbance}
\end{equation}
Notice that the gravity disturbance $\delta g_{P}$ is not equivalent 
to the magnitude of the gravity disturbance vector 
$\delta \mitbf{g}_{P}$ \citep{barthelmes2013, sanso_sideris2013}.
As properly pointed out by \citet{hackney-featherstone2003},
the gravity disturbance is a very-well established quantity in geodesy,
but appears to be less well known in geophysics.

%Consider a line (dashed line in Figure \ref{fig:surfaces}) passing 
%through a point $Q$ on the surface of the reference ellipsoid and 
%a point $P$ on the Earth's surface

PAREI AQUI - precisa descrever as demais quantidade na fig 3

@@

\begin{figure}
%    \vspace{5.5cm}
    \includegraphics{figures/cartesian-geodetic-systems.png}
    \caption{Schematic representation of the Geocentric Cartesian System (GCS) and the Geocentric Geodetic System (GGS). The GCS has the $Z$-axis coincident with the mean Earth's rotational axis,
    the $X$-axis pointing to the Greenwich meridian and the $Y$-axis
    directed so as to obtain a right-handed system. The GGS is      defined by an oblate ellipsoid with semi-minor axis $b$, 
    coincident with the $Z$-axis of GCS, and a semi-major
    axis $a$. In this coordinate system, the position of a point is
    determined by the geometric height $h$, geodetic latitude 
    $\varphi$ and longitude $\lambda$. 
    The Earth's center of mass is represented 
    by $O$, $P$ represents a point $(h_{P}, \varphi_{P}, \lambda_{P})$ and $P^{\prime}$ its projection onto the plane $XY$ 
    (Equatorial plane). The plane containing $O$, $P$ and 
    $P^{\prime}$ is represented in gray in (a) and (b). 
    The unit vectors $\hat{\mitbf{u}}_{P}$, $\hat{\mitbf{v}}_{P}$ and 
    $\hat{\mitbf{w}}_{P}$ define three mutually orthogonal 
    directions at $P$ (equation \ref{eq:unit-vectors}).
    In (b), $Q$ represents a point $(h_{Q}, \varphi_{Q}, \lambda_{Q})$,
    which is the projection of $P$ onto the reference ellipsoid, at
    the same latitude and longitude ($\varphi_{Q} = \varphi_{P}$ and $\lambda_{Q} = \lambda_{P}$).}
  \label{fig:GCS-GGS}
\end{figure}

\begin{figure}
    \includegraphics{figures/local-system.png}
    \caption{(a) Schematic representation of the gravity vector
    $\mitbf{g}_{P}$, normal gravity vector $\mitbf{\gamma}_{P}$,
    gravity disturbance vector $\delta \mitbf{g}_{P}$
    (equation \ref{eq:gravity-vector}) and unit vector 
    $\hat{\mitbf{u}}_{P}$ (equation \ref{eq:unit-vectors}) at a point 
    $P$.
    (b) Schematic representation of a topocentric Cartesian 
    coordinate system with origin at a point $P$. The axes $x$
    and $y$ are parallel to the unit vectors $\hat{\mitbf{v}}_{P}$
    and $\hat{\mitbf{w}}_{P}$ (equation \ref{eq:unit-vectors} and
    Figure \ref{fig:GCS-GGS}), respectively. On the other hand, the $z$ axis is opposite to the unit vector $\hat{\mitbf{u}}_{P}$ (equation 
    \ref{eq:unit-vectors} and Figure \ref{fig:GCS-GGS}) and points downward. The gray plane shown in (b) is the same gray plane
    shown in Figure \ref{fig:GCS-GGS}.}
  \label{fig:TCS}
\end{figure}

\begin{figure}
    \includegraphics{figures/surfaces.png}
    \caption{Schematic representation of the gravity vector
        $\mitbf{g}_{P}$, normal gravity vector $\mitbf{\gamma}_{P}$,
        gravity disturbance vector $\delta \mitbf{g}_{P}$
        (equation \ref{eq:gravity-vector}) and unit vector 
        $\hat{\mitbf{u}}_{P}$ (equation \ref{eq:unit-vectors}) at a point $P$ on the surface of the Earth; gravity vector 
        $\mitbf{g}_{Q^{\prime}}$ at a point $Q^{\prime}$ on the Geoid;
        normal gravity vector $\mitbf{\gamma}_{Q}$ at a point $Q$ on
        the reference ellipsoid. The dashed line is normal to the 
        surface of the reference ellipsoid at $Q$.}
  \label{fig:surfaces}
\end{figure}
